Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-11-01 response_to_reviewers.Rmd

> 11, 50: Given how much fivethirtyeight.com uses Elo ratings for all kinds of sports, it is probably worth mentioning their work as examples.

We have added the sentence: "Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at FiveThirtyEight.com." But again, we acknowledge that Wiley policy may not allow such a reference.

> 12, 1: As noted above, Section 4 feels much weaker than Sections 2 and 3 which are strong.  Perhaps a look at Pythagorean methods as a historical introduction to this topic.

We thank the reviewer for this excellent idea. We have added a new subsection on Expected Winning Percentages models to the beginning of Section 4.

@@

> 12, 23: This is not much of a case study.  Perhaps some additional information on the data available, ways it could be used to calculate game state expected valued, win probability, player ratings….

We have expanded the case study in chess to include Elo ratings...

In addition, we have added new subsection with a case study in baseball about the evolution of data, tools, and statistical methodologies. We hope that these additional will help to flesh out the value of these tools.

### 2022-11-01 sports_analytics.Rmd
added pythag; close #33

A third crucial idea in sports analytics is the estimation of team strength. 
A popular method for estimating team strength in sports is through pairwise evaluations.

@@

Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that
$$
\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j \,.
$$
Here, $\exp{\beta_i}$ is often viewed as a representation of team $i$'s ability.

>>>>

A third crucial idea in sports analytics is the estimation of team strength. 
First, we briefly introduce a simple method for estimating team strength based on won-loss record. 
Next, we detail three other more sophisticated methods for estimating team strength in sports through pairwise evaluations.

The impetus for all methods for estimating team strength is the realization that won-loss records are a noisy measure of team strength. 
As binary outcomes, and with all sports (except perhaps chess) involving some element of chance, wins and losses carry some signal of team strength, but we can do better.

## Expected winning percentage

The first method for estimating team strength to become popular in sports analytics in a measure of expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
Later, @miller2007derivation derived the formula as a consequence of assuming that runs are generated by two independent Weibull processes.

Expected winning percentage is just:
$$
  \widehat{wpct} = \frac{X^\alpha}{X^\alpha + Y^\alpha} \,,
$$
where $X$ is the number of points (runs) that a team has scored, and $Y$ is the number of points (runs) that they have allowed, over some specified time period. 
James's work was originally in baseball, and he posited the value of $\alpha = 2$. 
The resemblance to the formula for computing the length of the hypotenuse in a right triangle provides the nod to Pythagoras.

Subsequent analysts have tried to find the optimal value of $\alpha$ for various time periods. 
This can be done with a few lines of code, after observing that 
$$
  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^2}
$$
and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball.

```{r}
library(tidyverse)
library(Lahman)
exp_wpct <- function(x, k = 2) { 
  return(1/(1 + (1/x)^k))
}

TeamRuns <- Teams %>% 
  filter(yearID >= 1954) %>%
  rename(RS = R) %>% 
  mutate(WPct = W / (W + L), run_ratio = RS/RA) %>%
  select(yearID, teamID, lgID, WPct, run_ratio)

data_space <- ggplot(
  data = TeamRuns, aes(x = run_ratio, y = WPct)
) +
  geom_vline(xintercept = 1, color = "darkgray", linetype = 2) +
  geom_hline(yintercept = 0.5, color = "darkgray", linetype = 2) +
  geom_point(alpha = 0.2) + 
  xlab("Ratio of Runs Scored to Runs Allowed") + 
  ylab("Winning Percentage")
```

```{r}
optimal <- TeamRuns %>%
  nls(
    formula = WPct ~ 1/(1 + (1/run_ratio)^k), 
    start = list(k = 2)
  ) %>%
  coef()
```

```{r pythag, fig.cap="Winning percentages vs. runs scored and runs allowed in baseball, 1954--2021. The navy line represents the expected winning percentage model posited by James, with the exponent value of 2. The gold line shows the same model with an optimal exponent of 1.84."}
data_space +
  stat_function(fun = exp_wpct, args = list(2), size = 2, color = "navy") +
  stat_function(fun = exp_wpct, args = list(optimal), size = 2, color = "gold")
```

Many authors have fit expected winning percentage models to other sports---too many to cite here. 
See, for example, @hamilton2011extension for association football (soccer), @caro2013testing for Division I college football, and notably, future NBA general manager Daryl Morey for basketball [@dewan1993stats].

@@

Perhaps the most widely-used probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that
$$
\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j \,.
$$
Here, $\exp{\beta_i}$ is often viewed as a representation of team $i$'s ability.
